---
title: "Chapter 00: Title"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    social: menu
    source_code: embed
    theme: lumen
---

<style>
.desktop-layout {
  padding: 30px;
}
</style>

```{r}
# options for DT::datatable
dt_format <- list(
  searching = FALSE,
  paging = FALSE,
  scrollX = TRUE,
  scrollY = "300px")
```

Packages {data-navmenu="Explore" data-icon="fa-book"}
==================================================

## Packages

You will load the following packages in order to work with the code provided in this chapter.

```{r echo=TRUE}

```

```{r}
library(DT)
```


Definitions {data-navmenu="Explore" data-icon="fa-puzzle-piece"}
==================================================


Code examples {data-navmenu="Explore" data-icon="fa-code"}
==================================================


Tidy code examples {data-navmenu="Explore" data-icon="fa-magic"}
==================================================


```{r eval=FALSE}
# how to use the DT::datatable function
datatable(tidy_model, options = dt_format)
```

Extend {data-navmenu="Explore" data-icon="fa-flask"}
==================================================


